# Bash Programming

The last two chapters have discussed how to use the bash shell. Bash itself is a
little programming langauge, and this chapter we're going to discuss how you can
write your own computer programs in Bash. Programming in Bash is sueful to know
because of how seamlessly it integreats with all of the command line programs
you've already learned, and

## Math

## Variables

## User Input

## Arrays

## Braces

## If/Else

## Loops

## Functions

## Writing Programs

### Exercises

Write a program that takes one number as an argument and prints
"the number is even" if the number is even, "the number is odd" if the number
is odd, and "the number is less than 1" if the number is less than 1.

```{r, engine='bash', eval=FALSE}
evenodd 6
```

```
## the number is even
```

```{r, engine='bash', eval=FALSE}
evenodd 19
```

```
## the number is odd
```

```{r, engine='bash', eval=FALSE}
evenodd 0
```

```
## the number is less than 1
```

Write a program that takes many numbers as arguments and prints the largest
number. Add a flag to specify whether it prints the minimum or maximum number.

```{r, engine='bash', eval=FALSE}
findn -max 8 2 9 4 0 3
```

```
## 9
```

Write a program that takes one number as an argument and prints all of the
numbers
