# Git and GitHub

> Proof rather than argument. - Japanese proverb

## What are Git and GitHub?

Git is a command line program which allows you to track versions of any code or
plain text documents that you create. Like the "track changes" feature of
a word processor Git keeps track of who made particular changes, the time and 
date of those changes, and where the changes were made. If a critical file gets
deleted by accident, or if you make a breaking change to your code and you want
to try to figure out where the breaking change was made, you can use Git to
restore the deleted file or find the new bug in your program. Git organizes
groups of files that you're tracking into a **repository**, which is just
directory where all of the changes to files in that directory are tracked.

GitHub is a website that provides remote Git repositories. If you're working
on code together with a friend GitHub can help you sync changes to code files
between you and your friend. There's also a social and community aspect to 
GitHub, since you can watch other programmers develop their projects. GitHub 
also makes it easy to jump in and help somebody with their project. GitHub
offers many other useful features which will discuss at length.

## Setting Up Git and GitHub

Before setting up Git, go to [GitHub](https://github.com/) and create a free
account. Take note of which email address you use and which username you choose.

To see if you have Git installed open up your terminal and enter the following:

```{r, engine='bash'}
git --version
```

If you don't get a response back telling you the version of Git that you have
installed then you need to install Git. You can find instructions for installing
Git on your operating system 
[here](https://git-scm.com/book/en/v2/Getting-Started-Installing-Git).

Open up your shell once you have git installed and run `git --version` again to
make sure that installation succeeded (you may need to restart your shell or
your computer). After Git is installed we need to set up two environmental
variables, but we only need to do this once.


## Getting Started with Git

Let's create our first Git repository. First we need to create a directory:

```{r, engine='bash', eval=FALSE}
cd
mkdir my-first-repo
cd my-first-repo
```

"Repo" in this case is just shorthand for "repository." To start tracking files
with Git in a directory enter `git init` into the command line:

```{r, engine='bash', eval=FALSE}
git init
```

```
## Initialized empty Git repository in /Users/sean/my-first-repo/.git/
```

You've just created your first repository! Now let's create a file and start
tracking it.

```{r, engine='bash', eval=FALSE}
echo "Welcome to My First Repo" > readme.txt
```

Now that we've created a file in this Git repository, let's use `git status` to
see what's going on in this repository. We'll be using `git status` continuously
throughout this chapter in order to get information about the status Git repository.

```{r, engine='bash', eval=FALSE}
git status
```

```
## On branch master
## 
## Initial commit
## 
## Untracked files:
##   (use "git add <file>..." to include in what will be committed)
## 
## 	readme.txt
## 
## nothing added to commit but untracked files present (use "git add" to track)
```

As you can see `readme.txt` is listed as an untracked file. In order to let Git
know that you want to track this file we need to use `git add` with the name of
the file that we want to track. Let's start tracking `readme.txt`:

```{r, engine='bash', eval=FALSE}
git add readme.txt
```

Git now knows to track any changes to `readme.txt` let's see how the status of
the repository has changed:

```{r, engine='bash', eval=FALSE}
git status
```

```
## On branch master
## 
## Initial commit
## 
## Changes to be committed:
##   (use "git rm --cached <file>..." to unstage)
## 
## 	new file:   readme.txt
## 
```

Git is now tracking `readme.txt`, or in Git-specific language `readme.txt` is
now **staged**. Between the parentheses in the message above you can see that
`git status` is giving us a tip about how to unstage (or un-track) this file,
which we could do with `git rm --cached readme.txt`. Let's unstage this file
just to see what happens:

```{r, engine='bash', eval=FALSE}
git rm --cached readme.txt
```

```
## rm 'readme.txt'
```

```{r, engine='bash', eval=FALSE}
git status
```

```
## On branch master
## 
## Initial commit
## 
## Untracked files:
##   (use "git add <file>..." to include in what will be committed)
## 
## 	readme.txt
## 
```

Our repository is right back to the way it started with `readme.txt` as an
unstaged file. Let's start tracking `readme.txt` again so we can move on to
cooler Git features.

```{r, engine='bash', eval=FALSE}
git add readme.txt
```

Now that Git is tracking `readme.txt` we need to create a milestone to indicate
the changes that we made to `readme.txt`. In this case, the changes that we made
were creating the file in the first place! This milestone is called a **commit**
in Git. A commit logs the content of all of the currently staged files. Right
now we only have `readme.txt` staged so let's commit the creation of this file.
When making a Git commit, we need to write a commit message which is specified
after the `-m` flag. The message should breifly describe what changes you've
made since the last commit.

```{r, engine='bash', eval=FALSE}
git commit -m "added readme.txt"
```

```
## [master (root-commit) 73e53ca] added readme.txt
##  1 file changed, 1 insertion(+)
##  create mode 100644 readme.txt
```

The message above confirms that the commit succeeded and it summarizes the
changes that took place since the last commit. As you can see in the message we
only changed one file, and we only changed one line in that file. Let's run
`git status` again to see the state of our repository after we've made the first
commit:

```{r, engine='bash', eval=FALSE}
git status
```

```
## On branch master
## nothing to commit, working tree clean
```

All of the changes to the files in this repository have been committed! Let's
add a few more files to this repository and commit them.

```{r, engine='bash', eval=FALSE}
touch file1.txt
touch fil2.txt
ls
```

```
## file1.txt
## fil2.txt
## readme.txt
```

While we're at it let's also add a new line of text to `readme.txt`:

```{r, engine='bash', eval=FALSE}
echo "Learning Git is going well so far." >> readme.txt
```

Now that's we've added two more files and we've made changes to one file let's
take a look at the state of this repository.

```{r, engine='bash', eval=FALSE}
git status
```

```
## On branch master
## Changes not staged for commit:
##   (use "git add <file>..." to update what will be committed)
##   (use "git checkout -- <file>..." to discard changes in working directory)
## 
## 	modified:   readme.txt
## 
## Untracked files:
##   (use "git add <file>..." to include in what will be committed)
## 
## 	fil2.txt
## 	file1.txt
## 
## no changes added to commit (use "git add" and/or "git commit -a")
```

We can see that Git has detected that one file has been modified, and that there
are two files in this directory that it is not tracking. Now we needd to tell
Git to track the changes to these files. We could tell Git to track changes to
each file using `git add`, or since all of the files in this repository are
`.txt` files we could use a wildcard and enter `git add *.txt` into the
console. However if we want to track all of the changes to all of the files in
our directory we shoudl use the command `git add -A`.

```{r, engine='bash', eval=FALSE}
git add -A
git status
```

```
## On branch master
## Changes to be committed:
##   (use "git reset HEAD <file>..." to unstage)
## 
## 	new file:   fil2.txt
## 	new file:   file1.txt
## 	modified:   readme.txt
## 
```

Now the changes to all of the files in this repository are being tracked.
Finally let's commit these changes:

```{r, engine='bash', eval=FALSE}
git commit -m "added two files"
```

```
## [master 53a1983] added two files
##  3 files changed, 1 insertion(+)
##  create mode 100644 fil2.txt
##  create mode 100644 file1.txt
```

Darn it, now looking at this commit summary I realize that I have a typo in one
of the names of my files! Thankfully we can undo the most recent commit with 
the command `git reset --soft HEAD~`:

```{r, engine='bash', eval=FALSE}
git reset --soft HEAD~
git status
```

```
## On branch master
## Changes to be committed:
##   (use "git reset HEAD <file>..." to unstage)
## 
## 	new file:   fil2.txt
## 	new file:   file1.txt
## 	modified:   readme.txt
## 
```

This repo is now in that exact same state it was right before we made the 
commit. Now we can rename `fil2.txt` to `file2.txt`, then let's look at the
status of the repository again.

```{r, engine='bash', eval=FALSE}
mv fil2.txt file2.txt
git status
```

```
## On branch master
## Changes to be committed:
##   (use "git reset HEAD <file>..." to unstage)
## 
## 	new file:   fil2.txt
## 	new file:   file1.txt
## 	modified:   readme.txt
## 
## Changes not staged for commit:
##   (use "git add/rm <file>..." to update what will be committed)
##   (use "git checkout -- <file>..." to discard changes in working directory)
## 
## 	deleted:    fil2.txt
## 
## Untracked files:
##   (use "git add <file>..." to include in what will be committed)
## 
## 	file2.txt
## 
```

We previously told Git to track `fil2.txt`, and we can see that Git acknowledges
that the file has been deleted. We can bring Git up to speed with what files it
should be tracking with `git add -A`:

```{r, engine='bash', eval=FALSE}
git add -A
git status
```

```
## On branch master
## Changes to be committed:
##   (use "git reset HEAD <file>..." to unstage)
## 
## 	new file:   file1.txt
## 	new file:   file2.txt
## 	modified:   readme.txt
## 
```

Finally we got the file names right! Now let's make the correct commit:

```{r, engine='bash', eval=FALSE}
git commit -m "added two files"
```

```
## [master 12bb9f5] added two files
##  3 files changed, 1 insertion(+)
##  create mode 100644 file1.txt
##  create mode 100644 file2.txt
```

That looks much better.
